Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-30 paper.tex
Clear up some TODOs

\subsection{Calculating with the genotype matrix}
Storing genetic variation data compactly is important, but it is also
important that we can analyse the data efficiently. Bioinformatics 
workflows tend to emphasise text files and command line utilities 
that consume and produce text~\citep[e.g.][]{buffalo2015bioinformatics}. 
Thus, many tools that compress VCF data provide a command line 
utility with a query language to restrict the records
examined, perform some pre-specified calculations and finally 
output some text, typically VCF or tab/comma separated values
[TODO check these citations]~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
These pre-defined calculations are by necessity limited in scope, however,
and the volumes of text involved in Biobank scale datasets
make the classical approach of custom
analyses via Unix utilities in pipelines prohibitively slow. Thus, 
methods have begun to provide Application Programming Interfaces
(APIs), providing efficient access to genotype and other VCF data
[TODO check these citations]~\cite[e.g.][]{kelleher2013processing,lefaive2021sparse,
wertenbroek2022xsi,zhang2023gbc}. By providing programmatic access,
the data can be retrieved from storage, decoded and then analysed
in the same memory space without additional copies and 
inter-process communication through pipes.

@@

As datasets grow ever larger, the ability to efficiently access subsets 
of the data becomes increasingly important. VCF/BCF achieve efficient 
access to the data for genomic ranges 
by compressing blocks of adjacent records using \texttt{bgzip},
and storing secondary indexes alongside the original 
files with a conventional suffix~\citep{li2011tabix}. 
Thus, for a given range query we 
decompress only the necessary blocks and can quickly access
the required records. 
Methods like Savvy and XSI largely follow the same principles [TODO check and 
be more concrete]. The row-wise nature of these methods, however, mean 
that we cannot efficiently subset \emph{by sample}
(e.g., to calculate statistics within a particular cohort). In the extreme
case, if we want to access only the genotypes for a single sample
we must still retrieve and decompress the entire dataset.

>>>>

\subsection{Calculating with the genotype matrix}
Storing genetic variation data compactly is important, but it is also
important that we can analyse the data efficiently. Bioinformatics 
workflows tend to emphasise text files and command line utilities 
that consume and produce text~\citep[e.g.][]{buffalo2015bioinformatics}. 
Thus, many tools that compress VCF data provide a command line 
utility with a query language to restrict the records
examined, perform some pre-specified calculations and finally 
output some text, typically VCF or tab/comma separated 
values~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
zhang2023gbc}. %GBC
These pre-defined calculations are by necessity limited in scope, however,
and the volumes of text involved in Biobank scale datasets
make the classical approach of custom
analyses via Unix utilities in pipelines prohibitively slow. Thus, 
methods have begun to provide Application Programming Interfaces
(APIs), providing efficient access to genotype and other VCF 
data~\cite[e.g.][]{kelleher2013processing,lefaive2021sparse,
wertenbroek2022xsi}. By providing programmatic access,
the data can be retrieved from storage, decoded and then analysed
in the same memory space without additional copies and 
inter-process communication through pipes.

@@

As datasets grow ever larger, the ability to efficiently access subsets 
of the data becomes increasingly important. VCF/BCF achieve efficient 
access to the data for genomic ranges 
by compressing blocks of adjacent records using \texttt{bgzip},
and storing secondary indexes alongside the original 
files with a conventional suffix~\citep{li2011tabix}. 
Thus, for a given range query we 
decompress only the necessary blocks and can quickly access
the required records. 
The row-wise nature of VCF (and most proposed alternatives), however, means
that we cannot efficiently subset \emph{by sample}
(e.g., to calculate statistics within a particular cohort). In the extreme
case, if we want to access only the genotypes for a single sample
we must still retrieve and decompress the entire dataset.
